Scheduling and Amortization for MPC

### 2022-02-04 paper_SIMD.tex
Changed biometric example to literal code.
Edits to 4.5 Extension to Basic Vectorization.

\subsection{Example: Biometric}

@@

@@

The update $\texttt{A}[i,j]$ can be nested into an inner loop and there may be multiple updates, i.e., writes to $\texttt{A}[i,j]$. However, update such as $\texttt{A}[i-1,j] = ...$ or $\texttt{A}[i-1,j-1] = ...$, etc., is not allowed. Additionally, while there could be several different loops that perform canonical updates, they must be of the same dimensionality, i.e., an update of higher or lower dimension, e.g., $A[i,j,k] = ...$ is not allowed. We assign the \emph{canonical dimensionality} to each array in the obvious way. This restriction simplifies reasoning in this early stage of the compiler; we will look to relax the restriction in future work.

Reads through an arbitrary formula, such as $\texttt{A}[i-1]$ for example, are allowed, however, we assume the programmer ensures that the formula is within the bounds of the array.

@@

One change to Basic vectorization is the expansion of dimension if the array write or read occurs in a nested loop. That is, if there is an update $\texttt{A}[i,j] = ...$ that occurs in loop nest $i,j,k$, $\texttt{A}[i,j]$ will be rewritten into $\texttt{A}[i,j,k]$. Similarly, a read $\texttt{A}[f(i,j)]$ will be rewritten into $\texttt{A}[f(i,j),k]$. \ana{TODO: This notion of $f(i,j)$ needs a precise definition. For now I'm hoping the hand waving works.} This can be a preprocessing step or it can be incorporated into the Basic Vectorization algorithm.

The other change concerns def-use edges $X \rightarrow Y$ where $X$ defines and $Y$ uses an array variable (e.g., the definition can be an update $A_2 = update(A_1,...)$ or a pseudo $\phi$-node $A_2 = \phi(A_0,A_1)$). These edges are not handled in the same way as in Basic Vectorization, specifically, we do not expand dimension as we do for scalars in Basic Vectorization according to the loop nest. A key invariant is that the dimension of an array $A$ cannot drop below $A$'s canonical dimensionality. Below we enumerate the cases of def-use edges.

@@

\item inner-to-outer $X \rightarrow Y$. Let $A$ be the array defined at $X$. If the dimensionality of $A$ is greater than its canonical dimensionality, then add \emph{drop\_dim(...)} at $Y$, as in Basic Vectorization. Otherwise, do nothing.

\item outer-to-inner $X \rightarrow Y$. Add \emph{raise\_dim(...)} (at $X$) as in Basic Vectorization.

@@

After Phase 1 of Basic Vectorization the Aiken's array write example will be (roughly) as follows:

\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{A}_1$ = $\phi(\texttt{A}_0,\texttt{A}_2)$
\STATE 2. $\texttt{B}_1$ = $\phi(\texttt{B}_0,\texttt{B}_2)$
\STATE 3. $\texttt{C}_1$ = $\phi(\texttt{C}_0,\texttt{C}_2)$
\STATE 4. $\texttt{D}_2$ = $\phi(\texttt{D}_0,\texttt{D}_2)$
\STATE 5. $\texttt{A}_2 = update(\texttt{A}_1, \stackrel{\rightarrow}{i}, \texttt{B}_1[\stackrel{\rightarrow}{i}] + 10)$;
\STATE 6. $\texttt{B}_2 = update(\texttt{B}_1, \stackrel{\rightarrow}{i}, \texttt{A}_2[\stackrel{\rightarrow}{i}] * \texttt{D}_1[\stackrel{\rightarrow}{i-1}])$;
\STATE 7. $\texttt{C}_2 = update(\texttt{C}_1, \stackrel{\rightarrow}{i}, \texttt{A}_2[\stackrel{\rightarrow}{i}] * \texttt{D}_1[\stackrel{\rightarrow}{i-1}])$;
\STATE 8. $\texttt{D}_2 = update(\texttt{D}_1; \stackrel{\rightarrow}{i}, \texttt{B}_2[\stackrel{\rightarrow}{i}] * \texttt{C}_2[\stackrel{\rightarrow}{i}])$;
\end{algorithmic}

There are no nested loops, thus, array accesses remain as is. Phase 2 computes the closure of 4; $cl = \{4,6,7, 8\}$ 
while 5 is vectorizable. Recall that 1,2, and 3 are targetless phi-nodes. Since the closure $cl$ includes updates to $B$ and $C$,
the corresponding phi-nodes are added to the closure, i.e., the FOR loop. The uses of $A_1$ and $B_1$ in the vectorized statement
turn into uses of $A_0$ and $B_0$ respectively. (But note that $A_0$ is irrelevant; the updates writes into array $A_2$ in parallel.)

@@

\paragraph{Example 2}

@@

\begin{algorithmic}
\STATE ...
\STATE !2!3[$\stackrel{\rightarrow}i,\stackrel{\rightarrow}j$] = $\mathit{EQ\_SIMD}$(A[$\stackrel{\rightarrow}i,\stackrel{\rightarrow}j$], [[0,0...], [1,1...],...[num\_bins-1,num\_bins-1...]])
%\STATE \hspace{0.25cm}        result!4 = $\phi$(result!2, result!5)
\STATE FOR j=0; j<N; j++; \COMMENT{MOTION loop}
\STATE \hspace{0.25cm}            res!5 = $\phi$(res!4, res!7) \COMMENT{Do nothing, just copy.}
\STATE \hspace{0.25cm}            !3!3[$\stackrel{\rightarrow}i,j$] = $\mathit{ADD\_SIMD}$(res!5[$\stackrel{\rightarrow}i,j$],B[$\stackrel{\rightarrow}i,j$])
\STATE \hspace{0.25cm}            res!6 = res!5 \COMMENT{Do nothing, just copy.}
\STATE \hspace{0.25cm}            res!6[$\stackrel{\rightarrow}i,j$] = !3!3[$\stackrel{\rightarrow}i,j$]
\STATE \hspace{0.25cm}            res!7 = res!6
\STATE \hspace{0.25cm}            res!7[$\stackrel{\rightarrow}i,j$] = $\mathit{MUX\_SIMD}$(!2!3[$\stackrel{\rightarrow}i,j$], res!6[$\stackrel{\rightarrow}i,j$], res!5[$\stackrel{\rightarrow}i,j$])
\end{algorithmic}

>>>>

\input{biometric_vectorization_example.tex}

\begin{comment}
\subsection{Example: Biometric}

@@

\end{algorithmic}
\end{comment}

@@

The update $\texttt{A}[i,j]$ can be nested into an inner loop and there may be multiple updates, i.e., writes to $\texttt{A}[i,j]$. However, update such as $\texttt{A}[i-1,j] = ...$ or $\texttt{A}[i-1,j-1] = ...$, etc., is not allowed. Additionally, while there could be several different loops that perform canonical updates, they must be of the same dimensionality, i.e., an update of higher or lower dimension, e.g., $A[i,j,k] = ...$ is not allowed. We compute the \emph{canonical dimensionality} of each write array in the obvious way \emph{before basic vectorization}. This restriction simplifies reasoning in this early stage of the compiler; we will look to relax the restriction in future work.

Another restriction/assumption is that we assume the output array is given as input with initial values, and it is of size consistent with its canonical dimensionality. \ana{This restriction will require that we update our benchmarks that come with output arrays.}

Reads through an arbitrary formula, such as $\texttt{A}[i-1]$ for example, are allowed; currently, our projection function returns dummy values if the read formula is out of bounds; we assume the programmer ensures that the program still computes correct output in this case.

@@

One change to Basic vectorization is the expansion of dimension if the array write or read occurs in a nested loop. That is, if there is an update $\texttt{A}[i,j] = ...$ that occurs in loop dimensionality $i,j,k$, $\texttt{A}[i,j]$ will be rewritten into $\texttt{A}[i,j,k]$. Similarly, a read $\texttt{A}[f(i,j)]$ will be rewritten into $\texttt{A}[f(i,j),k]$. \ana{TODO: This notion of $f(i,j)$ needs a precise definition. For now I'm hoping the hand waving works.} This can be a preprocessing step or it can be incorporated into the Basic Vectorization algorithm.

The other change concerns def-use edges $X \rightarrow Y$ where $X$ defines and $Y$ uses an array variable, e.g., the definition can be an update $A_2 = update(A_1,...)$ or a pseudo $\phi$-node $A_2 = \phi(A_0,A_1)$. Note that $\phi$ nodes for arrays are no subscript operations on those $\phi$-nodes the way there are in scalar arrays. These edges are not handled in the same way as in Basic Vectorization, specifically, we do not raise and drop dimension as we do for scalars in Basic Vectorization. A key invariant is that the dimension of an array $A$ cannot go above or drop below $A$'s canonical dimensionality. We enumerate the cases of def-use edges.

@@

\item inner-to-outer $X \rightarrow Y$. If dimensionality of the loop enclosure of $X$ is greater than the canonical dimensionality of the array, then add \emph{drop\_dim(...)} at $Y$, as in Basic Vectorization. Otherwise, do nothing.

\item outer-to-inner $X \rightarrow Y$. If dimensionality of loop enclosure of $Y$ is greater than the canonical dimensionality of the array, then add \emph{raise\_dim(...)} (at $X$) as in Basic Vectorization. Otherwise, do nothing.

@@

First, the canonical dimensionality of all \texttt{A,B,C} and \texttt{D} is 1. 
Thus, there is no addition of extra dimension for inner loops. After Phase 1 
of Basic Vectorization the Aiken's array write example will be (roughly) as follows:

\begin{algorithmic}
\STATE for i in range(N):
\STATE 1. $\texttt{A}_1$ = $\phi(\texttt{A}_0,\texttt{A}_2)$
\STATE 2. $\texttt{B}_1$ = $\phi(\texttt{B}_0,\texttt{B}_2)$
\STATE 3. $\texttt{C}_1$ = $\phi(\texttt{C}_0,\texttt{C}_2)$
\STATE 4. $\texttt{D}_1$ = $\phi(\texttt{D}_0,\texttt{D}_2)$
\STATE 5. $\texttt{A}_2 = update(\texttt{A}_1, \stackrel{\rightarrow}{i}, \texttt{B}_1[\stackrel{\rightarrow}{i}] + 10)$;
\STATE 6. $\texttt{B}_2 = update(\texttt{B}_1, \stackrel{\rightarrow}{i}, \texttt{A}_2[\stackrel{\rightarrow}{i}] * \texttt{D}_1[\stackrel{\rightarrow}{i-1}])$;
\STATE 7. $\texttt{C}_2 = update(\texttt{C}_1, \stackrel{\rightarrow}{i}, \texttt{A}_2[\stackrel{\rightarrow}{i}] * \texttt{D}_1[\stackrel{\rightarrow}{i-1}])$;
\STATE 8. $\texttt{D}_2 = update(\texttt{D}_1; \stackrel{\rightarrow}{i}, \texttt{B}_2[\stackrel{\rightarrow}{i}] * \texttt{C}_2[\stackrel{\rightarrow}{i}])$;
\end{algorithmic}

Phase 2 computes the closure of 4; $cl = \{4,6,7, 8\}$ 
while 5 is vectorizable. Recall that 1,2, and 3 are targetless phi-nodes. Since the closure $cl$ includes updates to $B$ and $C$,
the corresponding phi-nodes are added to the closure, i.e., the FOR loop. The uses of $A_1$ and $B_1$ in the vectorized statement
turn into uses of $A_0$ and $B_0$ respectively. (But note that $A_0$ is irrelevant; the updates writes into array $A_2$ in parallel.)

@@

\ana{TODO: Format either Histogram or Matrix Multiplication. Both are interesting examples that use output arrays.}

\begin{comment}
\paragraph{Example 2}

@@

\begin{algorithmic}
\STATE ...
\STATE !2!3[$\stackrel{\rightarrow}i,\stackrel{\rightarrow}j$] = $\mathit{EQ\_SIMD}$(A[$\stackrel{\rightarrow}i,\stackrel{\rightarrow}j$], [[0,0...], [1,1...],...[num\_bins-1,num\_bins-1...]])
%\STATE \hspace{0.25cm}        result!4 = $\phi$(result!2, result!5)
\STATE FOR j=0; j<N; j++; \COMMENT{MOTION loop}
\STATE \hspace{0.25cm}            res!5 = $\phi$(res!4, res!7) \COMMENT{Do nothing, just copy.}
\STATE \hspace{0.25cm}            !3!3[$\stackrel{\rightarrow}i,j$] = $\mathit{ADD\_SIMD}$(res!5[$\stackrel{\rightarrow}i,j$],B[$\stackrel{\rightarrow}i,j$])
\STATE \hspace{0.25cm}            res!6 = res!5 \COMMENT{Do nothing, just copy.}
\STATE \hspace{0.25cm}            res!6[$\stackrel{\rightarrow}i,j$] = !3!3[$\stackrel{\rightarrow}i,j$]
\STATE \hspace{0.25cm}            res!7 = res!6
\STATE \hspace{0.25cm}            res!7[$\stackrel{\rightarrow}i,j$] = $\mathit{MUX\_SIMD}$(!2!3[$\stackrel{\rightarrow}i,j$], res!6[$\stackrel{\rightarrow}i,j$], res!5[$\stackrel{\rightarrow}i,j$])
\end{algorithmic}
\end{comment}
